---
title: "DEJU playback study data analysis"
author: "Eleanor, Kobe, Liam, Dave"
format: html
editor: source
---

```{r}
#| label: library_setup
#| include: false

library(tidyverse)
library(knitr)
library(hms)
```

# Data Import and Joining

Firstly, create a vector of all files in the data folder of the github repository which includes a `readme.md` file.

Then select only the `.txt` files to import into R.

```{r}
#| label: Data_File_Selection

data_files = list.files("data")

data_files = data_files[str_detect(data_files, "\\.txt")]
```

A lot of things happen in the next code chunk, to make it easier to follow the code chunk does the following:

1. Creates an empty data frame (tibble) that will eventually contain the full data set.
2. Begins a for loop to iterate through each file in the file list.
3. Reads in a data file, labeling the columns correctly.
4. Calculates the duration of each annotation in the `.txt` file from the onset and offset times.
5. Adds the file name from which each row came from to each of the rows of the data frame.
6. Separates the file name into the relevant information.
7. Fixes the formatting of the dates and times.
8. Creates a column of estimated distances
9. Adds the rows into the master data set.

```{r}
#| label: Create_Master_Data_Set

# 1
deju_data = tibble()

# 2
for(i in 1:length(data_files)) {
  
# 3
  dejuraw = as_tibble(read.table(paste0("data/", data_files[i]), sep = "\t", quote = "", col.names = c("Onset", "Offset", "Annotation")))
  
# 4
  deju_rtj = dejuraw |>
  mutate(Duration = Offset - Onset) |>
  relocate(Duration, .after = Offset) |>
# 5
  mutate(File_Name = data_files[i]) |>
  mutate(File_Name = str_remove(File_Name, "\\.txt")) |>
# 6
  separate(col = File_Name, into = c("Recording_Date", "Recording_Start_Time", "Bands", "Treatment"), sep = "_") |>
# 7
  mutate(Recording_Date = ymd(Recording_Date)) |>
  mutate(Recording_Start_Time = str_replace(Recording_Start_Time, "(\\d\\d)(\\d\\d)", "\\1:\\2")) |>
  mutate(Datetime = as.POSIXct(paste(Recording_Date, Recording_Start_Time, sep = " " ))) |>
  relocate(Datetime, .before = Recording_Start_Time) |>
# 8
  mutate(
    Estimated_Distance_of_Focal_Male = if_else(
      str_detect(Annotation, "^\\d+m$"),
      str_remove(Annotation, "m$"),
      NA_character_
    )
  ) |>
  fill(Estimated_Distance_of_Focal_Male, .direction = "down")|>
  mutate(Estimated_Distance_of_Focal_Male = as.numeric(Estimated_Distance_of_Focal_Male))|>
  select(-Recording_Start_Time)
  
# 9
  deju_data = bind_rows(deju_data, deju_rtj)
}
```

# Data Cleaning

Number of Songs...
```{r}
Songs = deju_data |>
  filter(str_detect(Annotation, "Focalsong")) |>
  group_by(Bands, Treatment) |>
  summarise(Number_Of_Songs = n(), .groups = "drop")
```

Minimum Distance...
```{r}
Minimum_Distance = deju_data |>
  group_by(Bands, Treatment) |>
  summarise(Minimum_Distance = if(all(is.na(Estimated_Distance_of_Focal_Male))) {
    NA_real_
  } else {
    min(Estimated_Distance_of_Focal_Male, na.rm = TRUE)
  }, .groups = "drop"
  )
```

Trouble shooting to figure out the conditional metrics

```{r}
Time_to_10m = deju_data |>
  group_by(Bands, Treatment) |>
  summarise(
    Pbstart_Time = first(Onset[Annotation == "Pbstart"]),
    First_Time_Within_10 = if_else(any(Estimated_Distance_of_Focal_Male <= 10, na.rm = TRUE),
     first(Onset[Estimated_Distance_of_Focal_Male <= 10], na_rm = TRUE),
     NA_real_),
    Time_To_Approach_10m = First_Time_Within_10 - Pbstart_Time,
    .groups = "drop"
  )
```

time spent within 10m

```{r}
deju_data |>
  group_by(Bands, Treatment) |>
  transform(Previous_Distance = lag(Estimated_Distance_of_Focal_Male)) |>
  transform(Next_Distance = lead(Estimated_Distance_of_Focal_Male)) |>
  mutate(Move = if_else(Estimated_Distance_of_Focal_Male < 10 & (Previous_Distance > 10 | is.na(Previous_Distance)) | Estimated_Distance_of_Focal_Male > 10 & Previous_Distance < 10, "yes", "no")) |>
  filter(Move == "yes") |>
  transform(Time_Spent = Onset - lag(Onset)) |>
  summarise(sum(Time_Spent, na.rm = TRUE))
```

## Eleanor lag/lead example
```{r}

deju_data %>% 
  filter(str_detect(Annotation, "^\\d+m$")) %>% 
  transform(next_distance = lead(Estimated_Distance_of_Focal_Male, n = 1)) %>%
  transform(prev_distance = lag(Estimated_Distance_of_Focal_Male, n = 1)) %>% 
  mutate(move = if_else(Estimated_Distance_of_Focal_Male < 10 & prev_distance > 10 | Estimated_Distance_of_Focal_Male > 10 & prev_distance < 10, "yes", "no")) %>% 
  filter(move == "yes") %>% 
  transform(time_spent = Onset - lag(Onset, n = 1)) %>% 
  summarize(sum(time_spent, na.rm = TRUE))
```

