---
title: "DEJU Playback Study Data Analysis"
author: "Eleanor, Kobe, Liam, Dave"
format:
  html:
    embed-resources: true
editor: source
---

# Setup

Load the packages used for data wrangling, date-time handling, summaries, and plotting.

```{r}
#| label: setup_libraries
#| include: false

library(tidyverse)
library(knitr)
library(hms)
library(see)
```

# Build the Annotation-Level Dataset

## Identify Annotation Files

List the `.txt` annotation files in the repositories `data` directory.

```{r}
#| label: data_file_selection

data_files = list.files("data")
data_files = data_files[str_detect(data_files, "\\.txt$")]
```

## Import and Combine Annotation Files

Import each annotation file, calculate annotation duration, extract trial metadata, from the file name, and carry forward the most recent focal-male distance estimate.

```{r}
#| label: build_annotation_dataset

# Create an empty data frame (tibble) that will eventually contain the full data set.
deju_data = tibble()

# Begin a for loop to iterate through each file in the file list.
for(i in 1:length(data_files)) {
  
# Read in a data file, labeling the columns correctly.
  dejuraw = as_tibble(
    read.table(
      paste0("data/", data_files[i]),
      sep = "\t",
      quote = "",
      col.names = c("onset", "offset", "annotation")
      )
    )
  
  deju_rtj = dejuraw %>%
# Calculate annotation duration
    mutate(duration = offset - onset) %>%
    relocate(duration, .after = offset) %>%
# Add the source file name and extract trial metadata.
    mutate(file_name = str_remove(data_files[i], "\\.txt$")) %>%
    separate_wider_delim(
      col = file_name,
      delim = "_",
      names = c("recording_date", "recording_start_time", "bands", "treatment")) %>%
# Format trial date and start time.
    mutate(recording_date = ymd(recording_date),
      recording_start_time = str_replace(
        recording_start_time,
        "(\\d\\d)(\\d\\d)",
        "\\1:\\2"),
      datetime = as.POSIXct(
        paste(recording_date, recording_start_time),
        tz = "America/St_Johns"
        )
      ) %>%
    relocate(datetime, .before = recording_start_time) %>%
# Extract and carry forward focal-male distance estimates.
    mutate(
      estimated_distance_of_focal_male = if_else(
        str_detect(annotation, "^\\d+m$"),
        str_remove(annotation, "m$"),
        NA_character_
      )
    ) %>%
    fill(estimated_distance_of_focal_male, .direction = "down") %>%
    mutate(estimated_distance_of_focal_male = as.numeric(estimated_distance_of_focal_male)) %>%
    select(-recording_start_time)
  
  deju_data = bind_rows(deju_data, deju_rtj)
}
```

# Derive Trial-Level Variables

Each section below reduces the annotation-level data to one response or trial descriptor per bird per treatment.

## Trial Date and Time

Extract the date and start time associated with each trial.

```{r}
#| label: summarise_trial_date_and_time

date_and_time = deju_data %>%
  group_by(bands, treatment) %>%
  summarise(
    date_of_trial = first(as_date(datetime)),
    time_of_trial = first(format(datetime, "%H:%M:%S")),
    .groups = "drop") %>%
  mutate(time_of_trial = as_hms(time_of_trial))
  
```

## Number of Songs

Count focal-male song annotations in each trial.

```{r}
#| label: summarise_number_of_songs

songs = deju_data %>%
  filter(str_detect(annotation, "Focalsong")) %>%
  group_by(bands, treatment) %>%
  summarise(number_of_songs = n(), .groups = "drop")
```

## Minimum Distance

Calculate the closest recorded distance to the playback speaker in each trial. Trials without a distance estimate remain `NA`.

```{r}
#| label: summarise_minimum_distance

minimum_distance = deju_data %>%
  group_by(bands, treatment) %>%
  summarise(
    minimum_distance = if(all(is.na(estimated_distance_of_focal_male))) {
      NA_real_
  } else {
    min(estimated_distance_of_focal_male, na.rm = TRUE)
  },
  .groups = "drop"
  )
```

## Time to Approach Within 10 Metres

Calculate elapsed time from playback start to the first recorded distance at or within 10 metres.

```{r}
#| label: summarise_time_to_10_metres

time_to_10m = deju_data %>%
  group_by(bands, treatment) %>%
  summarise(
    pbstart_time = first(onset[annotation == "Pbstart"]),
    first_time_within_10 = if_else(any(estimated_distance_of_focal_male <= 10, na.rm = TRUE),
     first(onset[estimated_distance_of_focal_male <= 10], na_rm = TRUE),
     NA_real_),
    time_to_approach_10m = first_time_within_10 - pbstart_time,
    .groups = "drop"
  )
```

## Time Spent Within 10 Metres

Estimate the total time spent within 10 metres from successive crossings of the 10 metre boundary

```{r}
#| label: summarise_time_within_ten_metres

time_within_10_metres = deju_data %>%
  group_by(bands, treatment) %>%
  mutate(
    previous_distance = lag(estimated_distance_of_focal_male),
    next_distance = lead(estimated_distance_of_focal_male),
    move = if_else(
      estimated_distance_of_focal_male <= 10 &
        (previous_distance > 10 | is.na(previous_distance)) |
        estimated_distance_of_focal_male > 10 & previous_distance <= 10,
      "yes",
      "no"
      )
    ) %>%
  filter(move == "yes") %>%
  mutate(
    direction = if_else(
      previous_distance > estimated_distance_of_focal_male,
      "in",
      "out"
      ),
    time_spent = if_else(direction == "in", NA_real_, onset - lag(onset))
    ) %>%
  summarise(
    time_within_10_metres = sum(time_spent, na.rm = TRUE),
    .groups = "drop"
    )
```

## Number of Songs Within 10 Metres

Count focal-male song annotations made while the current distance estimate is less than 10 metres.

```{r}
#| label: summarise_number_of_songs_within_10_metres

number_of_songs_within_10_metres = deju_data %>%
  group_by(bands, treatment) %>%
  filter(estimated_distance_of_focal_male < 10) %>%
  filter(str_detect(annotation, "Focalsong")) %>%
  summarise(
    number_of_songs_while_within_10_meters = n(),
    .groups = "drop"
    )
```

## Number of Physical Attacks

Count flyover annotations as physical attacks in each trial.

```{r}
#| label: summarise_physical_attacks

physical_attacks = deju_data %>%
  filter(str_detect(annotation, "Flyover")) %>%
  group_by(bands, treatment) %>%
  summarise(physical_attacks = n(), .groups = "drop")
```

## Number of Playback Stimuli

Count playback-stimulus annotations in each trial.

```{r}
#| label: summarise_stimulus_count

number_of_stimuli = deju_data %>%
  group_by(bands, treatment) %>%
  filter(str_detect(annotation, "Pbstim")) %>%
  summarise(number_of_stimuli = n(), .groups = "drop")
```

# Prepare Weather Covariates

## Import and Clean Weather Data

Combine the airport weather files, standardize time stamps, and retain temperature and wind speed for the current analysis data set.

```{r}
#| label: import_weather_data

file_list = list.files("SJIA_weather_data")

airport_weather_data = as_tibble(
  read.csv(paste0("SJIA_weather_data/", file_list[1]))
)

for(i in 2:length(file_list)) {
  temp_weather_data = as_tibble(
    read.csv(paste0("SJIA_weather_data/", file_list[i]))
  )
  
  airport_weather_data = bind_rows(
    airport_weather_data,
    temp_weather_data
  )
}

short_weather_data = airport_weather_data %>%
  mutate(datetime = as_datetime(Date.Time..LST., format = "%Y-%m-%d %H:%M"),
         datetime = force_tz(datetime, tzone = "America/St_Johns")) %>%
  transmute(
    datetime,
    temperature = Temp...C.,
    station_pressure = Stn.Press..kPa.,
    barometric_pressure = station_pressure * ((1 - (0.0065 * 140.50) / (temperature + 273.15 + (0.0065 * 140.50)))^-5.257),
    wind_speed = Wind.Spd..km.h.,
    wind_chill = Wind.Chill,
    relative_humidity = Rel.Hum....,
    dew_point = Dew.Point.Temp...C.,
  ) %>% 
  select(datetime, temperature, wind_speed)
```

# Assemble the Trial-Level Analysis Dataset

## Combine Trial Summaries

Create one row for every expected bird-treatment combination and join the trial-level response variables.

```{r}
#| label: build_trial_dataset

dark_eyed_juncos = c("SRBB", "SGGR", "SBYY", "SYYR", "SBYB", "SRBY", "SGGG", "SYBB", "SBBB", "SRBR", "SRRG")

conditions = c("Control", "10", "16", "22")

temp = c()

for(i in dark_eyed_juncos) {
  for(j in conditions) {
    temp = c(temp, paste0(i, "_", j))
  }
}

cleaned_summary_dataset = tibble(bird_condition = temp) %>%
  separate(col = bird_condition, into = c("bands", "treatment")) %>%
  left_join(date_and_time, by = c("bands", "treatment")) %>%
  left_join(songs, by = c("bands", "treatment")) %>%
  left_join(minimum_distance, by = c("bands", "treatment")) %>%
  left_join(time_to_10m, by = c("bands", "treatment")) %>%
  left_join(time_within_10_metres, by = c("bands", "treatment")) %>%
  left_join(number_of_songs_within_10_metres, by = c("bands", "treatment")) %>%
  left_join(physical_attacks, by = c("bands", "treatment")) %>%
  left_join(number_of_stimuli, by = c("bands", "treatment")) %>%
  mutate(
    treatment = recode_factor(
      treatment,
      "Control" = "Control",
      "22" = "One SD Below Mean",
      "16" = "Mean",
      "10" = "One SD Above Mean",
      .ordered = TRUE
    )
  ) %>%
  relocate(date_of_trial, time_of_trial, .before = bands) %>%
  select(-c(pbstart_time, first_time_within_10))
```

## Joining weather variables to associated trials

Match each trial to the weather observation at the correspoonding half-hour time stamp.

```{r}
#| label: join_weather_covariates

cleaned_summary_dataset = cleaned_summary_dataset %>% 
  filter(!is.na(date_of_trial)) %>% 
  mutate(
    datetime = ymd_hms(paste(date_of_trial, time_of_trial)),
    datetime = force_tz(datetime, tzone = "America/St_Johns"),
    closest_datetime = floor_date(datetime, unit = "hour") + minutes(30)
  ) %>%
  left_join(
    short_weather_data,
    by = c("closest_datetime" = "datetime")
  ) %>% 
  select(-ends_with(".y"), -ends_with(".x"))
```

::: {.callout-note}
## Checks to Resolve Before Formal Modeling

- Whether the summaries are correctly summarising each of the variables (10% Check?)
- Whether `time_to_approach_10m` should explicitly ignore distance observations before `Pbstart`.
- Whether "within 10 metres" should consistently mean `< 10` or `<= 10` across variables.
- Whether trials with zero songs, attacks, or songs within 10 metres should be recorded as `0` rather than `NA` after the joins
- wehter the `time_within_10_metres` calculation correctly handles trials that begin or end while the bird is within 10 metres.
-Whether the half-hour weather time stamp is the intended nearest weather observation for every trial.
:::

# Exploratory Analysis

## Number of Songs

::: {.panel-tabset}

### Overall Histogram

```{r}
#| label: song_count_histogram

ggplot(cleaned_summary_dataset, aes(x = number_of_songs, fill = treatment)) +
  geom_histogram(colour = "black") +
  labs(
    x = "Number of Songs",
    y = "Count",
    fill = "Treatment"
  ) +
  theme_bw() +
  scale_fill_manual(values = c("gray50", "#1b9e77", "#d95f02", "#7570b3"))
```

```{r}
#| label: song_count_maximum
#| include: false

max(cleaned_summary_dataset$number_of_songs, na.rm = TRUE)
```

### Mean and Standard Deviation

```{r}
#| label: song_count_mean_sd

ggplot(
  cleaned_summary_dataset,
  aes(x = treatment, y = number_of_songs, fill = treatment)
) +
  stat_summary(fun = mean, geom = "col", colour = "black") +
  stat_summary(
    fun.data = mean_sdl,
    fun.args = list(mult = 1),
    geom = "errorbar",
    width = 0.2
  ) +
  labs(
    x = "Treatment",
    y = "Number of Songs",
    fill = "Treatment"
  ) +
  scale_fill_manual(values = c("gray50", "#1b9e77", "#d95f02", "#7570b3"))+
  theme_bw() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(face = "bold")
  )
```

### Median and Median Absolute Deviation

```{r}
#| label: song_count_median_mad

ggplot(
  cleaned_summary_dataset,
  aes(x = treatment, y = number_of_songs, fill = treatment)
) +
  stat_summary(fun = median, geom = "col", colour = "black") +
  stat_summary(
    fun.data = function(x) {
      med = median(x, na.rm = TRUE)
      mad_val = mad(x, na.rm = TRUE)
    
      tibble(
        y = med,
        ymin = med - mad_val,
        ymax = med + mad_val
      )
    },
    geom = "errorbar",
    width = 0.2
  ) +
  labs(
    x = "Treatment",
    y = "Median Number of Songs",
    fill = "Treatment"
  ) +
  scale_fill_manual(values = c("gray50", "#1b9e77", "#d95f02", "#7570b3")) +
  theme_bw() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(face = "bold")
  )
```

### Summary Statistics

```{r}
#| label: song_count_summary

numbers_for_summary_bar_charts = cleaned_summary_dataset %>%
  group_by(treatment) %>%
  summarise(
    total_songs = sum(number_of_songs, na.rm = TRUE),
    mean = mean(number_of_songs, na.rm = TRUE),
    standard_deviation = sd(number_of_songs, na.rm = TRUE),
    median = median(number_of_songs, na.rm = TRUE),
    median_absolute_deviation = mad(number_of_songs, na.rm = TRUE),
    .groups = "drop"
  )

glimpse(numbers_for_summary_bar_charts)
```

### Distribution by treatment

```{r}
#| label: song_count_distribution

ggplot(cleaned_summary_dataset, aes(x = treatment,y = number_of_songs)) +
  geom_boxplot(
    aes(fill = treatment),
    width = 0.2,
    outlier.colour = "firebrick2"
  ) +
  geom_violinhalf(
    aes(fill = treatment),
    alpha = 0.25,
    position = position_nudge(x = 0.10)
  ) +
  geom_jitter(alpha = 0.5, width = 0.05) +
  scale_fill_manual(values = c("gray50", "#1b9e77", "#d95f02", "#7570b3")) +
  labs(
    x = "Treatment",
    y = "Number of Songs",
    fill = "Treatment"
  ) +
  theme_bw() +
  theme(axis.text.x = element_blank())
```

:::

## Individual Treatment Profiles

Plot each bird across treatments to visualize the repeated-measures structure.

```{r}
#| label: song_count_individual_profiles

ggplot(
  cleaned_summary_dataset,
  aes(
    x = treatment,
    y = number_of_songs,
    group = bands,
    color = bands
  )
) +
  geom_point() + 
  geom_line()
```

## Song-Count Sensitivity to Extreme Observations

The following data set excludes observations outside 1.5 IQR of the treatment specific distribution. It is retained for visualization and sensitivity checks only; exclusion from formal analyses should require a biological or measurement-based justification.

```{r}
#| label: filter_song_count_outliers

number_of_songs_no_outliers = cleaned_summary_dataset %>%
  group_by(treatment) %>%
  filter(
    number_of_songs >= quantile(number_of_songs, 0.25, na.rm = TRUE) - 1.5 * IQR(number_of_songs, na.rm = TRUE),
    number_of_songs <= quantile(number_of_songs, 0.75, na.rm = TRUE) + 1.5 * IQR(number_of_songs, na.rm = TRUE)
  ) %>%
  ungroup()
```

::: {.panel-tabset}

### Histogram Without Extreme Observations

```{r}
#| label: song_count_histogram_no_outliers

ggplot(number_of_songs_no_outliers, aes(x = number_of_songs, fill = treatment)) +
  geom_histogram(colour = "black") +
  labs(
    x = "Number of Songs",
    y = "Count",
    fill = "Treatment"
  ) +
  theme_bw() +
  scale_fill_manual(values = c("gray50", "#1b9e77", "#d95f02", "#7570b3"))
```

```{r}
#| label: maximum_number_of_songs_no_outliers
#| include: false

max(number_of_songs_no_outliers$number_of_songs, na.rm = TRUE)
```

### Mean and Standard Deviation Without Extreme Observations

```{r}
#| label: song_count_mean_sd_no_outliers

ggplot(
  number_of_songs_no_outliers,
  aes(x = treatment, y = number_of_songs, fill = treatment)
) +
  stat_summary(fun = mean, geom = "col", colour = "black") +
  stat_summary(
    fun.data = mean_sdl,
    fun.args = list(mult = 1),
    geom = "errorbar",
    width = 0.2
  ) +
  labs(
    x = "Treatment",
    y = "Number of Songs",
    fill = "Treatment"
  ) +
  scale_fill_manual(values = c("gray50", "#1b9e77", "#d95f02", "#7570b3"))+
  theme_bw() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(face = "bold")
  )
```

### Median and MAD Without Extreme Observations

```{r}
#| label: song_count_median_mad_no_outliers

ggplot(
  number_of_songs_no_outliers,
  aes(x = treatment, y = number_of_songs, fill = treatment)
) +
  stat_summary(fun = median, geom = "col", colour = "black") +
  stat_summary(
    fun.data = function(x) {
      med = median(x, na.rm = TRUE)
      mad_val = mad(x, na.rm = TRUE)
    
      tibble(
        y = med,
        ymin = med - mad_val,
        ymax = med + mad_val
      )
    },
    geom = "errorbar",
    width = 0.2
  ) +
  labs(
    x = "Treatment",
    y = "Median Number of Songs",
    fill = "Treatment"
  ) +
  scale_fill_manual(values = c("gray50", "#1b9e77", "#d95f02", "#7570b3")) +
  theme_bw() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(face = "bold"))
```

### Summary Statistics Without Extreme Observations

```{r}
#| label: song_count_summary_no_outliers

numbers_for_summary_bar_charts_no_outliers = number_of_songs_no_outliers %>%
  group_by(treatment) %>%
  summarise(
    total_songs = sum(number_of_songs, na.rm = TRUE),
    mean = mean(number_of_songs, na.rm = TRUE),
    standard_deviation = sd(number_of_songs, na.rm = TRUE),
    median = median(number_of_songs, na.rm = TRUE),
    median_absolute_deviation = mad(number_of_songs, na.rm = TRUE),
    .groups = "drop"
  )

glimpse(numbers_for_summary_bar_charts_no_outliers)
```

### Distribution Without Extreme Observations

```{r}
#| label: song_count_distribution_no_outliers

ggplot(number_of_songs_no_outliers, aes(x = treatment,y = number_of_songs)) +
  geom_boxplot(
    aes(fill = treatment),
    width = 0.2,
    outlier.colour = "firebrick2"
  ) +
  geom_violinhalf(
    aes(fill = treatment),
    alpha = 0.25,
    position = position_nudge(x = 0.10)
  ) +
  geom_jitter(alpha = 0.5, width = 0.05) +
  scale_fill_manual(values = c("gray50", "#1b9e77", "#d95f02", "#7570b3")) +
  labs(
    x = "Treatment",
    y = "Number of Songs",
    fill = "Treatment"
  ) +
  theme_bw() +
  theme(axis.text.x = element_blank())
```

:::

## Number of Songs While Within 10 Metres

::: {.panel-tabset}

### Overall Histogram

```{r}
#| label: songs_within_10_metres_histogram

ggplot(
  cleaned_summary_dataset,
  aes(x = number_of_songs_while_within_10_meters, fill = treatment)
) +
  geom_histogram(binwidth = 5, colour = "black") +
  labs(
    x = "Number of Songs While Within 10 Metres",
    y = "Count",
    fill = "Treatment"
  ) +
  scale_x_continuous(
    breaks = seq(
      0,
      ceiling(max(
        cleaned_summary_dataset$number_of_songs,
        na.rm = TRUE
        ) / 10) * 10,
      10
    )
  ) +
  theme_bw() +
  scale_fill_manual(values = c("gray50", "#1b9e77", "#d95f02", "#7570b3"))
```

```{r}
#| label: songs_within_10_metres_maximum
#| include: false

max(cleaned_summary_dataset$number_of_songs_while_within_10_meters, na.rm = TRUE)
```

### Total Songs by Treatment

```{r}
#| label: songs_within_10_metres_total_bar_chart

ggplot(
  cleaned_summary_dataset,
  aes(
    x = treatment,
    y = number_of_songs_while_within_10_meters,
    fill = treatment
  )
) +
  geom_col(colour = "black") +
  labs(
    x = "Treatment",
    y = "Number of Songs While Within 10 Metres",
    fill = "Treatment"
  ) +
  scale_fill_manual(values = c("gray50", "#1b9e77", "#d95f02", "#7570b3"))+
  theme_bw() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(face = "bold"))
```

### Treatment Totals

```{r}
#| label: songs_within_10_metres_treatment_totals

cleaned_summary_dataset %>%
  group_by(treatment) %>%
  summarise(
    total_songs_within_10_metres = sum(
      number_of_songs_while_within_10_meters,
      na.rm = TRUE
    ),
    .groups = "drop"
  )
```

### Distribution by Treatment

```{r}
#| label: songs_within_10_metres_distribution

ggplot(
  cleaned_summary_dataset,
  aes(x = treatment,y = number_of_songs_while_within_10_meters)
) +
  geom_boxplot(
    aes(fill = treatment),
    width = 0.2,
    outlier.colour = "firebrick2"
  ) +
  geom_violinhalf(
    aes(fill = treatment),
    alpha = 0.25,
    position = position_nudge(x = 0.10)
  ) +
  geom_jitter(alpha = 0.5, width = 0.05) +
  scale_fill_manual(values = c("gray50", "#1b9e77", "#d95f02", "#7570b3")) +
  labs(
    x = "Treatment",
    y = "Number of Songs While Within 10 Meteres",
    fill = "Treatment"
  ) +
  theme_bw() +
  theme(axis.text.x = element_blank())
```
:::

# Exploratory Song--Stimulus Overlap

```{r}
#| label: identify song_stimulus_overlap

overlap = deju_data %>%
  mutate(
    comparison_onset = if_else(
      str_detect(annotation, "Pbstim"),
      onset,
      NA_real_
    ),
    comparison_onset_reverse = comparison_onset,
    comparison_offset = if_else(
      str_detect(annotation, "Pbstim"),
      offset,
      NA_real_
    ),
    comparison_offset_reverse = comparison_offset
  ) %>%
  fill(comparison_onset, comparison_offset, .direction = "down") %>%
  fill(comparison_onset_reverse, comparison_offset_reverse, .direction = "up") %>%
  mutate(
    focalsong_onset = if_else(
      str_detect(annotation, "Focalsong"),
      onset,
      NA_real_
    ),
    focalsong_offset = if_else(
      str_detect(annotation, "Focalsong"),
      offset,
      NA_real_
    ),
    overlap = if_else(
      focalsong_onset >= comparison_onset &
        focalsong_onset <= comparison_offset,
      "overlap",
      NA_character_
    ),
    reverse_overlap = if_else(
      focalsong_offset >= comparison_onset_reverse &
        focalsong_offset <= comparison_offset_reverse,
      "overlap",
      NA_character_
    )
  ) 
```

based on what the results of the above code chunk it seems to look like DEJU intentionally do not overlap there songs. some more data analysis should be run to confirm this but preliminary results based off of some what was produced in the data frame suggest that Junco overlapping stimulus is ten times less likely than stimulus overlapping Junco.

