---
title: "DEJU playback study data analysis"
author: "Eleanor, Kobe, Liam, Dave"
format: html
editor: source
---

```{r}
#| label: library_setup
#| include: false

library(tidyverse)
library(knitr)
library(hms)
```

Initially, I will develop some code that creates the variables for one text file and then I will generalize it so that it will iterate through all files and calculate the variables automatically

## Data Import

Select only the `.txt` files in the data folder and assign them to a vector/list.
```{r}
data_files = list.files("data")

data_files = data_files[str_detect(data_files, "\\.txt")]
```

Actually import the data

```{r}
dejuraw = as_tibble(read.table(paste0("data/", data_files[1]), sep = "\t", quote = "", col.names = c("Onset", "Offset", "Annotation")))
```

Calculate duration of each annotation and relocate it cause its prettier and makes more sense.

```{r}
deju_rtj = dejuraw |>
  mutate(Duration = Offset - Onset) |>
  relocate(Duration, .after = Offset) |>
  mutate(File_Name = data_files[1]) |>
  mutate(File_Name = str_remove(File_Name, "\\.txt")) |>
  separate(col = File_Name, into = c("Recording_Date", "Recording_Start_Time", "Bands", "Treatment"), sep = "_") |>
  mutate(Recording_Date = ymd(Recording_Date)) |>
  mutate(Recording_Start_Time = str_replace(Recording_Start_Time, "(\\d\\d)(\\d\\d)", "\\1:\\2")) |>
  mutate(Datetime = as.POSIXct(paste(Recording_Date, Recording_Start_Time, sep = " " ))) |>
  relocate(Datetime, .before = Recording_Start_Time) |>
  select(-Recording_Start_Time)
```

