---
title: "DEJU Playback Study Data Analysis"
author: "Eleanor, Kobe, Liam, Dave"
format:
  html:
    embed-resources: true
editor: source
---

# Setup

Load the packages used for data wrangling, date-time handling, summaries, and plotting.

```{r}
#| label: setup_libraries
#| include: false

library(tidyverse)
library(knitr)
library(hms)
library(see)
```

# Build the Annotation-Level Dataset

## Identify Annotation Files

List the `.txt` annotation files in the repositories `data` directory.

```{r}
#| label: data_file_selection

data_files = list.files("data")
data_files = data_files[str_detect(data_files, "\\.txt$")]
```

## Import and Combine Annotation Files

Import each annotation file, calculate annotation duration, extract trial metadata, from the file name, and carry forward the most recent focal-male distance estimate.

```{r}
#| label: build_annotation_dataset

# Create an empty data frame (tibble) that will eventually contain the full data set.
deju_data = tibble()

# Begin a for loop to iterate through each file in the file list.
for(i in 1:length(data_files)) {
  
# Read in a data file, labeling the columns correctly.
  dejuraw = as_tibble(
    read.table(
      paste0("data/", data_files[i]),
      sep = "\t",
      quote = "",
      col.names = c("onset", "offset", "annotation")
      )
    )
  
  deju_rtj = dejuraw %>%
# Calculate annotation duration
    mutate(duration = offset - onset) %>%
    relocate(duration, .after = offset) %>%
# Add the source file name and extract trial metadata.
    mutate(file_name = str_remove(data_files[i], "\\.txt$")) %>%
    separate_wider_delim(
      col = file_name,
      delim = "_",
      names = c("recording_date", "recording_start_time", "bands", "treatment")) %>%
# Format trial date and start time.
    mutate(recording_date = ymd(recording_date),
      recording_start_time = str_replace(
        recording_start_time,
        "(\\d\\d)(\\d\\d)",
        "\\1:\\2"),
      datetime = as.POSIXct(
        paste(recording_date, recording_start_time),
        tz = "America/St_Johns"
        )
      ) %>%
    relocate(datetime, .before = recording_start_time) %>%
# Extract and carry forward focal-male distance estimates.
    mutate(
      estimated_distance_of_focal_male = if_else(
        str_detect(annotation, "^\\d+m$"),
        str_remove(annotation, "m$"),
        NA_character_
      )
    ) %>%
    fill(estimated_distance_of_focal_male, .direction = "down") %>%
    mutate(estimated_distance_of_focal_male = as.numeric(estimated_distance_of_focal_male)) %>%
    select(-recording_start_time)
  
  deju_data = bind_rows(deju_data, deju_rtj)
}
```

# Derive Trial-Level Variables

Each section below reduces the annotation-level data to one response or trial descriptor per bird per treatment.

## Trial Date and Time

Extract the date and start time associated with each trial.

```{r}
#| label: summarise_trial_date_and_time

date_and_time = deju_data %>%
  group_by(bands, treatment) %>%
  summarise(
    date_of_trial = first(as_date(datetime)),
    time_of_trial = first(format(datetime, "%H:%M:%S")),
    .groups = "drop") %>%
  mutate(time_of_trial = as_hms(time_of_trial))
  
```

## Number of Songs

Count focal-male song annotations in each trial.

```{r}
#| label: summarise_number_of_songs

songs = deju_data %>%
  filter(str_detect(annotation, "Focalsong")) %>%
  group_by(bands, treatment) %>%
  summarise(number_of_songs = n(), .groups = "drop")
```

## Minimum Distance

Calculate the closest recorded distance to the playback speaker in each trial. Trials without a distance estimate remain `NA`.

```{r}
#| label: summarise_minimum_distance

minimum_distance = deju_data %>%
  group_by(bands, treatment) %>%
  summarise(
    minimum_distance = if(all(is.na(estimated_distance_of_focal_male))) {
      NA_real_
  } else {
    min(estimated_distance_of_focal_male, na.rm = TRUE)
  },
  .groups = "drop"
  )
```

## Time to Approach Within 10 Metres

Calculate elapsed time from playback start to the first recorded distance at or within 10 metres.

```{r}
#| label: summarise_time_to_10_metres

time_to_10m = deju_data %>%
  group_by(bands, treatment) %>%
  summarise(
    pbstart_time = first(onset[annotation == "Pbstart"]),
    first_time_within_10 = if_else(any(estimated_distance_of_focal_male <= 10, na.rm = TRUE),
     first(onset[estimated_distance_of_focal_male <= 10], na_rm = TRUE),
     NA_real_),
    time_to_approach_10m = first_time_within_10 - pbstart_time,
    .groups = "drop"
  )
```

## Time Spent Within 10 Metres

Estimate the total time spent within 10 metres from successive crossings of the 10 metre boundary

```{r}
#| label: summarise_time_within_ten_metres

time_within_10_metres = deju_data %>%
  group_by(bands, treatment) %>%
  mutate(
    previous_distance = lag(estimated_distance_of_focal_male),
    next_distance = lead(estimated_distance_of_focal_male),
    move = if_else(
      estimated_distance_of_focal_male <= 10 &
        (previous_distance > 10 | is.na(previous_distance)) |
        estimated_distance_of_focal_male > 10 & previous_distance <= 10,
      "yes",
      "no"
      )
    ) %>%
  filter(move == "yes") %>%
  mutate(
    direction = if_else(
      previous_distance > estimated_distance_of_focal_male,
      "in",
      "out"
      ),
    time_spent = if_else(direction == "in", NA_real_, onset - lag(onset))
    ) %>%
  summarise(
    time_within_10_metres = sum(time_spent, na.rm = TRUE),
    .groups = "drop"
    )
```

## Number of Songs Within 10 Metres

Count focal-male song annotations made while the current distance estimate is less than 10 metres.

```{r}
#| label: summarise_number_of_songs_within_10_metres

number_of_songs_within_10_metres = deju_data %>%
  group_by(bands, treatment) %>%
  filter(estimated_distance_of_focal_male <= 10) %>%
  filter(str_detect(annotation, "Focalsong")) %>%
  summarise(
    number_of_songs_while_within_10_meters = n(),
    .groups = "drop"
    )
```

## Number of Physical Attacks

Count flyover annotations as physical attacks in each trial.

```{r}
#| label: summarise_physical_attacks

physical_attacks = deju_data %>%
  filter(str_detect(annotation, "Flyover")) %>%
  group_by(bands, treatment) %>%
  summarise(physical_attacks = n(), .groups = "drop")
```

## Number of Playback Stimuli

Count playback-stimulus annotations in each trial.

```{r}
#| label: summarise_stimulus_count

number_of_stimuli = deju_data %>%
  group_by(bands, treatment) %>%
  filter(str_detect(annotation, "Pbstim")) %>%
  summarise(number_of_stimuli = n(), .groups = "drop")
```

# Prepare Weather Covariates

## Import and Clean Weather Data

Combine the airport weather files, standardize time stamps, and retain temperature and wind speed for the current analysis data set.

```{r}
#| label: import_weather_data

file_list = list.files("SJIA_weather_data")

airport_weather_data = as_tibble(
  read.csv(paste0("SJIA_weather_data/", file_list[1]))
)

for(i in 2:length(file_list)) {
  temp_weather_data = as_tibble(
    read.csv(paste0("SJIA_weather_data/", file_list[i]))
  )
  
  airport_weather_data = bind_rows(
    airport_weather_data,
    temp_weather_data
  )
}

short_weather_data = airport_weather_data %>%
  mutate(datetime = as_datetime(Date.Time..LST., format = "%Y-%m-%d %H:%M"),
         datetime = force_tz(datetime, tzone = "America/St_Johns")) %>%
  transmute(
    datetime,
    temperature = Temp...C.,
    station_pressure = Stn.Press..kPa.,
    barometric_pressure = station_pressure * ((1 - (0.0065 * 140.50) / (temperature + 273.15 + (0.0065 * 140.50)))^-5.257),
    wind_speed = Wind.Spd..km.h.,
    wind_chill = Wind.Chill,
    relative_humidity = Rel.Hum....,
    dew_point = Dew.Point.Temp...C.,
  ) %>% 
  select(datetime, temperature, wind_speed)
```

# Assemble the Trial-Level Analysis Data Set

## Combine Trial Summaries

Create one row for every expected bird-treatment combination and join the trial-level response variables.

```{r}
#| label: build_trial_dataset

dark_eyed_juncos = c("SRBB", "SGGR", "SBYY", "SYYR", "SBYB", "SRBY", "SGGG", "SYBB", "SBBB", "SRBR", "SRRG")

conditions = c("Control", "10", "16", "22")

temp = c()

for(i in dark_eyed_juncos) {
  for(j in conditions) {
    temp = c(temp, paste0(i, "_", j))
  }
}

cleaned_summary_dataset = tibble(bird_condition = temp) %>%
  separate(col = bird_condition, into = c("bands", "treatment")) %>%
  left_join(date_and_time, by = c("bands", "treatment")) %>%
  left_join(songs, by = c("bands", "treatment")) %>%
  left_join(minimum_distance, by = c("bands", "treatment")) %>%
  left_join(time_to_10m, by = c("bands", "treatment")) %>%
  left_join(time_within_10_metres, by = c("bands", "treatment")) %>%
  left_join(number_of_songs_within_10_metres, by = c("bands", "treatment")) %>%
  left_join(physical_attacks, by = c("bands", "treatment")) %>%
  left_join(number_of_stimuli, by = c("bands", "treatment")) %>%
  mutate(
    treatment = recode_factor(
      treatment,
      "Control" = "Control",
      "22" = "One SD Below Mean",
      "16" = "Mean",
      "10" = "One SD Above Mean",
      .ordered = TRUE
    )
  ) %>%
  relocate(date_of_trial, time_of_trial, .before = bands) %>%
  select(-c(pbstart_time, first_time_within_10)) %>%
  mutate(
    time_within_10_metres = replace_na(time_within_10_metres, 0),
    number_of_songs = replace_na(number_of_songs, 0),
    number_of_songs_while_within_10_meters = replace_na(number_of_songs_while_within_10_meters, 0),
    physical_attacks = replace_na(physical_attacks, 0)
      )
```

## Joining weather variables to associated trials

Match each trial to the weather observation at the correspoonding half-hour time stamp.

```{r}
#| label: join_weather_covariates

cleaned_summary_dataset = cleaned_summary_dataset %>% 
  filter(!is.na(date_of_trial)) %>% 
  mutate(
    datetime = ymd_hms(paste(date_of_trial, time_of_trial)),
    datetime = force_tz(datetime, tzone = "America/St_Johns"),
    closest_datetime = floor_date(datetime, unit = "hour") + minutes(30)
  ) %>%
  left_join(
    short_weather_data,
    by = c("closest_datetime" = "datetime")
  ) %>% 
  select(-ends_with(".y"), -ends_with(".x"))
```

::: {.callout-note}
## Checks to Resolve Before Formal Modeling

- Validate the derived trial-level behavioural summaries against a subset of original annotations.
- Confirm that `time_to_approach_10m` ignores distance observations occurring before `Pbstart`.
- Standardize whether "within 10 metres" means `<= 10` throughout the analysis.
- Verify the handling of trials that begin or end while the focal male is within 10 metres when calculating `time_within_10_metres`.
- Confirm that the selected half-hour weather observation is the intended weather match for each trial.
- Validate automated song--stimulus overlap classifications against the original Audacity annotations.
- Recover calibrated playback amplitudes, if available, for the three conspecific amplitude treatments.
:::

# Exploratory Analysis

## Number of Songs

::: {.panel-tabset}

### Overall Histogram

```{r}
#| label: song_count_histogram

ggplot(cleaned_summary_dataset, aes(x = number_of_songs, fill = treatment)) +
  geom_histogram(colour = "black") +
  labs(
    x = "Number of Songs",
    y = "Count",
    fill = "Treatment"
  ) +
  theme_bw() +
  scale_fill_manual(values = c("gray50", "#1b9e77", "#d95f02", "#7570b3"))
```

```{r}
#| label: song_count_maximum
#| include: false

max(cleaned_summary_dataset$number_of_songs, na.rm = TRUE)
```

### Mean and Standard Deviation

```{r}
#| label: song_count_mean_sd

ggplot(
  cleaned_summary_dataset,
  aes(x = treatment, y = number_of_songs, fill = treatment)
) +
  stat_summary(fun = mean, geom = "col", colour = "black") +
  stat_summary(
    fun.data = mean_sdl,
    fun.args = list(mult = 1),
    geom = "errorbar",
    width = 0.2
  ) +
  labs(
    x = "Treatment",
    y = "Number of Songs",
    fill = "Treatment"
  ) +
  scale_fill_manual(values = c("gray50", "#1b9e77", "#d95f02", "#7570b3"))+
  theme_bw() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(face = "bold")
  )
```

### Median and Median Absolute Deviation

```{r}
#| label: song_count_median_mad

ggplot(
  cleaned_summary_dataset,
  aes(x = treatment, y = number_of_songs, fill = treatment)
) +
  stat_summary(fun = median, geom = "col", colour = "black") +
  stat_summary(
    fun.data = function(x) {
      med = median(x, na.rm = TRUE)
      mad_val = mad(x, na.rm = TRUE)
    
      tibble(
        y = med,
        ymin = med - mad_val,
        ymax = med + mad_val
      )
    },
    geom = "errorbar",
    width = 0.2
  ) +
  labs(
    x = "Treatment",
    y = "Median Number of Songs",
    fill = "Treatment"
  ) +
  scale_fill_manual(values = c("gray50", "#1b9e77", "#d95f02", "#7570b3")) +
  theme_bw() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(face = "bold")
  )
```

### Summary Statistics

```{r}
#| label: song_count_summary

numbers_for_summary_bar_charts = cleaned_summary_dataset %>%
  group_by(treatment) %>%
  summarise(
    total_songs = sum(number_of_songs, na.rm = TRUE),
    mean = mean(number_of_songs, na.rm = TRUE),
    standard_deviation = sd(number_of_songs, na.rm = TRUE),
    median = median(number_of_songs, na.rm = TRUE),
    median_absolute_deviation = mad(number_of_songs, na.rm = TRUE),
    .groups = "drop"
  )

glimpse(numbers_for_summary_bar_charts)
```

### Distribution by treatment

```{r}
#| label: song_count_distribution

ggplot(cleaned_summary_dataset, aes(x = treatment,y = number_of_songs)) +
  geom_boxplot(
    aes(fill = treatment),
    width = 0.2,
    outlier.colour = "firebrick2"
  ) +
  geom_violinhalf(
    aes(fill = treatment),
    alpha = 0.25,
    position = position_nudge(x = 0.10)
  ) +
  geom_jitter(alpha = 0.5, width = 0.05) +
  scale_fill_manual(values = c("gray50", "#1b9e77", "#d95f02", "#7570b3")) +
  labs(
    x = "Treatment",
    y = "Number of Songs",
    fill = "Treatment"
  ) +
  theme_bw() +
  theme(axis.text.x = element_blank())
```

:::

## Time within 10 metres

::: {.panel-tabset}

### Overall Histogram

```{r}
#| label: time_within_10_metres_histogram

ggplot(cleaned_summary_dataset, aes(x = time_within_10_metres, fill = treatment)) +
  geom_histogram(colour = "black") +
  labs(
    x = "Time within 10m",
    y = "Count",
    fill = "Treatment"
  ) +
  theme_bw() +
  scale_fill_manual(values = c("gray50", "#1b9e77", "#d95f02", "#7570b3"))
```

### Mean and Standard Deviation

```{r}
#| label: within_10_metres_mean_sd

ggplot(
  cleaned_summary_dataset,
  aes(x = treatment, y = time_within_10_metres, fill = treatment)
) +
  stat_summary(fun = mean, geom = "col", colour = "black") +
  stat_summary(
    fun.data = mean_sdl,
    fun.args = list(mult = 1),
    geom = "errorbar",
    width = 0.2
  ) +
  labs(
    x = "Treatment",
    y = "Time within 10m",
    fill = "Treatment"
  ) +
  scale_fill_manual(values = c("gray50", "#1b9e77", "#d95f02", "#7570b3"))+
  theme_bw() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(face = "bold")
  )
```

### Distribution by treatment

```{r}
#| label: within_10_metres_distribution

ggplot(cleaned_summary_dataset, aes(x = treatment,y = time_within_10_metres)) +
  geom_boxplot(
    aes(fill = treatment),
    width = 0.2,
    outlier.colour = "firebrick2"
  ) +
  geom_violinhalf(
    aes(fill = treatment),
    alpha = 0.25,
    position = position_nudge(x = 0.10)
  ) +
  geom_jitter(alpha = 0.5, width = 0.05) +
  scale_fill_manual(values = c("gray50", "#1b9e77", "#d95f02", "#7570b3")) +
  labs(
    x = "Treatment",
    y = "Time within 10m",
    fill = "Treatment"
  ) +
  theme_bw() +
  theme(axis.text.x = element_blank())
```

:::

## Time to approach 10 metres

::: {.panel-tabset}

### Overall Histogram

```{r}
#| label: time_to_approach_10m_histogram

ggplot(cleaned_summary_dataset, aes(x = time_to_approach_10m, fill = treatment)) +
  geom_histogram(colour = "black") +
  labs(
    x = "Time to approach 10m",
    y = "Count",
    fill = "Treatment"
  ) +
  theme_bw() +
  scale_fill_manual(values = c("gray50", "#1b9e77", "#d95f02", "#7570b3"))
```

:::

## Minimum distance

::: {.panel-tabset}

### Overall Histogram

```{r}
#| label: minimum_distance_histogram

ggplot(cleaned_summary_dataset, aes(x = minimum_distance, fill = treatment)) +
  geom_histogram(colour = "black") +
  labs(
    x = "Minimum distance (m)",
    y = "Count",
    fill = "Treatment"
  ) +
  theme_bw() +
  scale_fill_manual(values = c("gray50", "#1b9e77", "#d95f02", "#7570b3"))
```

### Mean and Standard Deviation

```{r}
#| label: minimum_distance_mean_sd

ggplot(
  cleaned_summary_dataset,
  aes(x = treatment, y = minimum_distance, fill = treatment)
) +
  stat_summary(fun = mean, geom = "col", colour = "black") +
  stat_summary(
    fun.data = mean_sdl,
    fun.args = list(mult = 1),
    geom = "errorbar",
    width = 0.2
  ) +
  labs(
    x = "Treatment",
    y = "Minimum distance (m)",
    fill = "Treatment"
  ) +
  scale_fill_manual(values = c("gray50", "#1b9e77", "#d95f02", "#7570b3"))+
  theme_bw() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(face = "bold")
  )
```

:::

## Individual Treatment Profiles

Plot each bird across treatments to visualize the repeated-measures structure.

```{r}
#| label: song_count_individual_profiles

ggplot(
  cleaned_summary_dataset,
  aes(
    x = treatment,
    y = number_of_songs,
    group = bands,
    color = bands
  )
) +
  geom_point() + 
  geom_line()
```

## Song-Count Sensitivity to Extreme Observations

The following data set excludes observations outside 1.5 IQR of the treatment specific distribution. It is retained for visualization and sensitivity checks only; exclusion from formal analyses should require a biological or measurement-based justification.

```{r}
#| label: filter_song_count_outliers

number_of_songs_no_outliers = cleaned_summary_dataset %>%
  group_by(treatment) %>%
  filter(
    number_of_songs >= quantile(number_of_songs, 0.25, na.rm = TRUE) - 1.5 * IQR(number_of_songs, na.rm = TRUE),
    number_of_songs <= quantile(number_of_songs, 0.75, na.rm = TRUE) + 1.5 * IQR(number_of_songs, na.rm = TRUE)
  ) %>%
  ungroup()
```

::: {.panel-tabset}

### Histogram Without Extreme Observations

```{r}
#| label: song_count_histogram_no_outliers

ggplot(number_of_songs_no_outliers, aes(x = number_of_songs, fill = treatment)) +
  geom_histogram(colour = "black") +
  labs(
    x = "Number of Songs",
    y = "Count",
    fill = "Treatment"
  ) +
  theme_bw() +
  scale_fill_manual(values = c("gray50", "#1b9e77", "#d95f02", "#7570b3"))
```

```{r}
#| label: maximum_number_of_songs_no_outliers
#| include: false

max(number_of_songs_no_outliers$number_of_songs, na.rm = TRUE)
```

### Mean and Standard Deviation Without Extreme Observations

```{r}
#| label: song_count_mean_sd_no_outliers

ggplot(
  number_of_songs_no_outliers,
  aes(x = treatment, y = number_of_songs, fill = treatment)
) +
  stat_summary(fun = mean, geom = "col", colour = "black") +
  stat_summary(
    fun.data = mean_sdl,
    fun.args = list(mult = 1),
    geom = "errorbar",
    width = 0.2
  ) +
  labs(
    x = "Treatment",
    y = "Number of Songs",
    fill = "Treatment"
  ) +
  scale_fill_manual(values = c("gray50", "#1b9e77", "#d95f02", "#7570b3"))+
  theme_bw() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(face = "bold")
  )
```

### Median and MAD Without Extreme Observations

```{r}
#| label: song_count_median_mad_no_outliers

ggplot(
  number_of_songs_no_outliers,
  aes(x = treatment, y = number_of_songs, fill = treatment)
) +
  stat_summary(fun = median, geom = "col", colour = "black") +
  stat_summary(
    fun.data = function(x) {
      med = median(x, na.rm = TRUE)
      mad_val = mad(x, na.rm = TRUE)
    
      tibble(
        y = med,
        ymin = med - mad_val,
        ymax = med + mad_val
      )
    },
    geom = "errorbar",
    width = 0.2
  ) +
  labs(
    x = "Treatment",
    y = "Median Number of Songs",
    fill = "Treatment"
  ) +
  scale_fill_manual(values = c("gray50", "#1b9e77", "#d95f02", "#7570b3")) +
  theme_bw() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(face = "bold"))
```

### Summary Statistics Without Extreme Observations

```{r}
#| label: song_count_summary_no_outliers

numbers_for_summary_bar_charts_no_outliers = number_of_songs_no_outliers %>%
  group_by(treatment) %>%
  summarise(
    total_songs = sum(number_of_songs, na.rm = TRUE),
    mean = mean(number_of_songs, na.rm = TRUE),
    standard_deviation = sd(number_of_songs, na.rm = TRUE),
    median = median(number_of_songs, na.rm = TRUE),
    median_absolute_deviation = mad(number_of_songs, na.rm = TRUE),
    .groups = "drop"
  )

glimpse(numbers_for_summary_bar_charts_no_outliers)
```

### Distribution Without Extreme Observations

```{r}
#| label: song_count_distribution_no_outliers

ggplot(number_of_songs_no_outliers, aes(x = treatment,y = number_of_songs)) +
  geom_boxplot(
    aes(fill = treatment),
    width = 0.2,
    outlier.colour = "firebrick2"
  ) +
  geom_violinhalf(
    aes(fill = treatment),
    alpha = 0.25,
    position = position_nudge(x = 0.10)
  ) +
  geom_jitter(alpha = 0.5, width = 0.05) +
  scale_fill_manual(values = c("gray50", "#1b9e77", "#d95f02", "#7570b3")) +
  labs(
    x = "Treatment",
    y = "Number of Songs",
    fill = "Treatment"
  ) +
  theme_bw() +
  theme(axis.text.x = element_blank())
```

:::

## Number of Songs While Within 10 Metres

::: {.panel-tabset}

### Overall Histogram

```{r}
#| label: songs_within_10_metres_histogram

ggplot(
  cleaned_summary_dataset,
  aes(x = number_of_songs_while_within_10_meters, fill = treatment)
) +
  geom_histogram(binwidth = 5, colour = "black") +
  labs(
    x = "Number of Songs While Within 10 Metres",
    y = "Count",
    fill = "Treatment"
  ) +
  scale_x_continuous(
    breaks = seq(
      0,
      ceiling(max(
        cleaned_summary_dataset$number_of_songs,
        na.rm = TRUE
        ) / 10) * 10,
      10
    )
  ) +
  theme_bw() +
  scale_fill_manual(values = c("gray50", "#1b9e77", "#d95f02", "#7570b3"))
```

```{r}
#| label: songs_within_10_metres_maximum
#| include: false

max(cleaned_summary_dataset$number_of_songs_while_within_10_meters, na.rm = TRUE)
```

### Total Songs by Treatment

```{r}
#| label: songs_within_10_metres_total_bar_chart

ggplot(
  cleaned_summary_dataset,
  aes(
    x = treatment,
    y = number_of_songs_while_within_10_meters,
    fill = treatment
  )
) +
  geom_col(colour = "black") +
  labs(
    x = "Treatment",
    y = "Number of Songs While Within 10 Metres",
    fill = "Treatment"
  ) +
  scale_fill_manual(values = c("gray50", "#1b9e77", "#d95f02", "#7570b3"))+
  theme_bw() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(face = "bold"))
```

### Treatment Totals

```{r}
#| label: songs_within_10_metres_treatment_totals

cleaned_summary_dataset %>%
  group_by(treatment) %>%
  summarise(
    total_songs_within_10_metres = sum(
      number_of_songs_while_within_10_meters,
      na.rm = TRUE
    ),
    .groups = "drop"
  )
```

### Distribution by Treatment

```{r}
#| label: songs_within_10_metres_distribution

ggplot(
  cleaned_summary_dataset,
  aes(x = treatment,y = number_of_songs_while_within_10_meters)
) +
  geom_boxplot(
    aes(fill = treatment),
    width = 0.2,
    outlier.colour = "firebrick2"
  ) +
  geom_violinhalf(
    aes(fill = treatment),
    alpha = 0.25,
    position = position_nudge(x = 0.10)
  ) +
  geom_jitter(alpha = 0.5, width = 0.05) +
  scale_fill_manual(values = c("gray50", "#1b9e77", "#d95f02", "#7570b3")) +
  labs(
    x = "Treatment",
    y = "Number of Songs While Within 10 Meteres",
    fill = "Treatment"
  ) +
  theme_bw() +
  theme(axis.text.x = element_blank())
```
:::

## Focal Song Duration

Summarise focal-male song duration to examine variation among playback treatments and individual birds.

### Mean Song Duration by Treatment

```{r}
#| label: summarise_focal_song_duration

focal_duration_by_treatment = deju_data %>%
  filter(str_detect(annotation, "Focalsong")) %>%
  mutate(
    treatment = recode_factor(
      treatment,
      "Control" = "Control",
      "22" = "One SD Below Mean",
      "16" = "Mean",
      "10" = "One SD Above Mean",
      .ordered = TRUE
    )
  ) %>%
  group_by(bands, treatment) %>%
  summarise(mean_song_duration = mean(duration),
            sd_song_duration = sd(duration),
            .groups = "drop")
```

### Individual Treatment Profiles

Plot the mean focal song duration for each bird across playback treatments.

```{r}
#| label: plot_focal_song_duration_by_treatment

ggplot(
  focal_duration_by_treatment,
  aes(x = treatment,
      y = mean_song_duration,
      group = bands,
      colour = bands
    )
) +
  geom_line() +
  geom_point() +
  labs(
    x = "Treatment",
    y = "Mean Song Duration (s)",
    colour = "Bands"
  ) +
  theme_bw()
```

### Bird Level Song Duration

Summarise focal-song duration across all treatments for each bird. standard errors are calculated from the number of songs produced by each bird.

```{r}
#| label: summarise_focal_song_duration_by_bird

focal_duration_by_bird = deju_data %>%
  filter(str_detect(annotation, "Focalsong")) %>%
  group_by(bands) %>%
  summarise(
    mean_song_duration = mean(duration),
    sd_song_duration = sd(duration),
    se_song_duration = sd_song_duration/(sqrt(n())),
    number_of_songs = n(),
    .groups = "drop"
  )
```

```{r}
#| label: plot_focal_song_duration_by_bird

ggplot(
  focal_duration_by_bird,
  aes(
    x = bands,
    y = mean_song_duration
  )
) +
  geom_point() +
  geom_errorbar(
    aes(
      ymin = mean_song_duration - se_song_duration,
      ymax = mean_song_duration + se_song_duration,
  ),
    width = 0.2
  ) +
  labs(
    x = "Bird",
    y = "Mean Song Duration (s)"
  ) +
  theme_bw()
```

# Song--Stimulus Overlap Analysis

## Analysis Rationale

Song--stimulus overlap was quantified following the general approach of Wilson et al. (2016). An overlap was defined as a focal-male song that began while a playback stimulus was already being broadcast. This directional definition evaluates whether focal males initiate songs during playback stimuli.

For each trial, the observed number of focal-song overlaps will be compared with the number expected if focal-song timing were independent of playback timing. Chance expectation is calculated from the playback duty cycle, defined as the proportion of the experimental playback period occupied by playback stimuli.

Because the playback experiment manipulated stimulus amplitude, focal songs used for the amplitude-related overlap analysis are restricted to songs produced while the focal male was estimated to be within 10 metres of the playback speaker. This restriction reduces variation in perceived playback amplitude caused by distance from the speaker.

## Define Playback Windows

Identify the beginning and end of the experimental playback period for each trial.

The beginning of the playback is identified by the `Pbstart` annotation. Most trials also contain a `Pbstop` annotation identifying the end of the playback. However, because `Pbstop` is missing from some trials, the offset of the final `Pbstim` annotation is used for the playback endpoint.

```{r}
#| label: define_playback_windows

playback_windows = deju_data %>%
  group_by(bands, treatment) %>%
  summarise(
    playback_start = first(onset[str_detect(annotation, "Pbstart")]),
    playback_end = last(offset[str_detect(annotation, "Pbstim")]),
    .groups = "drop"
  ) %>%
  mutate(playback_duration = playback_end - playback_start)
```

## Identify Playback Stimuli

Create one row for each playback stimulus. The annotation onset and offset define the interval during which the stimulus was playing.

```{r}
#| label: identify_playback_stimuli

playback_stimuli = deju_data %>%
  filter(str_detect(annotation, "Pbstim")) %>%
  group_by(bands, treatment) %>%
  arrange(onset, .by_group = TRUE) %>%
  transmute(
    bands,
    treatment,
    stimulus_number = row_number(),
    stimulus_onset = onset,
    stimulus_offset = offset,
    stimulus_duration = duration
  ) %>%
  ungroup()
```

## Identify Focal Songs

Create one row for each focal-male song and retain its onset, offset, duration, and estimated distance from the playback speaker.

```{r}
#| label: identify_focal_songs

focal_songs = deju_data %>%
  filter(str_detect(annotation, "Focalsong")) %>%
  group_by(bands, treatment) %>%
  arrange(onset, .by_group = TRUE) %>%
  transmute(
    bands,
    treatment,
    song_number = row_number(),
    song_onset = onset,
    song_offset = offset,
    song_duration = duration,
    estimated_distance_of_focal_male
  ) %>%
  ungroup()
```

## Define Eligible Focal Songs

Restrict the analysis to focal songs produced during the experimental playback period while the focal male was estimated to be within 10 metres of the playback speaker.

```{r}
#| label: define_eligible_focal_songs

eligible_focal_songs = focal_songs %>%
  left_join(playback_windows, by = c("bands", "treatment")) %>%
  filter(song_onset >= playback_start,
         song_onset <= playback_end,
         estimated_distance_of_focal_male <= 10)
```

## Classify Observed Overlaps

A focal song is classified as an overlap when its onset occurs after the beginning and before the end of a playback stimulus. Songs beginning outside all playback-stimulus intervals are classified as nonoverlaps.

```{r}
#| label: classify_song_overlap

song_overlap_matches = eligible_focal_songs %>%
  inner_join(
    playback_stimuli,
    by = join_by(
      bands,
      treatment,
      song_onset >= stimulus_onset,
      song_onset < stimulus_offset
    )
  ) %>%
  select(
    bands,
    treatment,
    song_number,
    stimulus_number
  )
```

```{r}
#| label: create_song_level_overlap_dataset

song_level_overlap = eligible_focal_songs %>%
  left_join(
    song_overlap_matches,
    by = c("bands", "treatment", "song_number")
  ) %>%
  mutate(
    overlap = if_else(
    is.na(stimulus_number),
    0L,
    1L
  )
)
```

## Calculate Playback Duty Cycle

Calculate the proportion of each experimental playback period occupied by playback stimuli. Duty cycle represents the probability that a focal song beginning at a randomly selected time would begin during a playback stimulus.

```{r}
#| label: calculate_playback_duty_cycle

playback_duty_cycle = playback_stimuli %>%
  group_by(bands, treatment) %>%
  summarise(
    total_stimulus_time = sum(stimulus_duration, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  left_join(
    playback_windows,
    by = c("bands", "treatment")
  ) %>%
  mutate(
    duty_cycle = total_stimulus_time / playback_duration
  )
```

## Summarise Observed Overlaps

For each bird-treatment trial, count the number of eligible focal songs, count the number that began during playback stimuli, and calculate the observed proportion of focal songs that overlapped playback.

```{r}
#| label: summarise_observed_overlaps

observed_overlaps = song_level_overlap %>%
  group_by(bands, treatment) %>%
  summarise(
    number_of_focal_songs = n(),
    observed_overlaps = sum(overlap),
    observed_overlap_proportion = mean(overlap),
    .groups = "drop"
  )
```

## Calculate Expected Overlap

Calculate the number of focal-song overlaps expected under random song timing. Expected overlap is the number of eligible focal songs multiplied by the playback duty cycle.

```{r}
#| label: calculate_expected_overlaps

trial_overlap_summary = playback_duty_cycle %>%
  left_join(
    observed_overlaps,
    by = c("bands", "treatment")
  ) %>%
  mutate(
    number_of_focal_songs = replace_na(
      number_of_focal_songs,
      0L
    ),
    observed_overlaps = replace_na(
      observed_overlaps,
      0L
    ),
    observed_overlap_proportion = if_else(
      number_of_focal_songs > 0,
      observed_overlaps / number_of_focal_songs,
      NA_real_
    ),
    expected_overlaps =
      number_of_focal_songs * duty_cycle,
    expected_overlap_proportion = duty_cycle,
    observed_expected_ratio = if_else(
      expected_overlaps > 0,
      observed_overlaps / expected_overlaps,
      NA_real_
    )
  )
```

## Inspect Observed and Expected Overlap

Summarise observed and expected overlap before fitting inferential models.

```{r}
#| label: summarise_observed_expected_overlap

trial_overlap_summary %>%
  filter(number_of_focal_songs > 0) %>%
  summarise(
    number_of_trials = n(),
    total_focal_songs = sum(number_of_focal_songs),
    total_observed_overlaps = sum(observed_overlaps),
    total_expected_overlaps = sum(expected_overlaps),
    median_observed_overlaps = median(observed_overlaps),
    median_expected_overlaps = median(expected_overlaps),
    median_observed_expected_ratio = median(
      observed_expected_ratio,
      na.rm = TRUE
    )
  )
```

```{r}
#| label: summarise_overlap_by_treatment

trial_overlap_summary %>%
  filter(number_of_focal_songs > 0) %>%
  group_by(treatment) %>%
  summarise(
    number_of_trials = n(),
    total_focal_songs = sum(number_of_focal_songs),
    total_observed_overlaps = sum(observed_overlaps),
    total_expected_overlaps = sum(expected_overlaps),
    median_observed_overlaps = median(observed_overlaps),
    median_expected_overlaps = median(expected_overlaps),
    median_observed_expected_ratio = median(
      observed_expected_ratio,
      na.rm = TRUE
    ),
    .groups = "drop"
  )
```

# Looking for covariation
In this section, we examine the possibility of correlation between response variables and the possibility of impacts of some random factors like weather

## Correlogram
Plotting all response variables + weather data against each other

```{r}
#| label: correlogram of response variables and weather data

library(ggforce)

ggplot(cleaned_summary_dataset) +
  geom_point(aes(x = .panel_x, y = .panel_y)) +
  facet_matrix(vars(wind_speed, temperature, number_of_songs, minimum_distance, time_to_approach_10m, time_within_10_metres, number_of_songs_while_within_10_meters))

```








